A non-contextual hidden variable model for quantum mechanics

### 2017-02-08 finalPaper.tex
/ ‘finalPaper.tex’ / ‘references.bib’

\documentclass[british,aps,prl,superscriptaddress,nofootinbib,times,reprint]{revtex4-1}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm,headheight=2cm,headsep=2cm,footskip=2cm}
\setcounter{secnumdepth}{3}
\usepackage{verbatim}
\usepackage{refstyle}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{esint}
\makeatletter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
\AtBeginDocument{\providecommand\thmref[1]{\ref{thm:#1}}}
\AtBeginDocument{\providecommand\lemref[1]{\ref{lem:#1}}}
\AtBeginDocument{\providecommand\propref[1]{\ref{prop:#1}}}
\AtBeginDocument{\providecommand\eqref[1]{\ref{eq:#1}}}
\AtBeginDocument{\providecommand\figref[1]{\ref{fig:#1}}}
\AtBeginDocument{\providecommand\secref[1]{\ref{sec:#1}}}
\AtBeginDocument{\providecommand\ssecref[1]{\ref{ssec:#1}}}
\AtBeginDocument{\providecommand\defnref[1]{\ref{defn:#1}}}
\AtBeginDocument{\providecommand\tblref[1]{\ref{tbl:#1}}}

@@

\begin{document}
\title{A non-contextual hidden variable model for quantum mechanics}
\author{Atul Singh Arora}
\email{ms11003@iisermohali.ac.in}
\selectlanguage{british}%
\affiliation{Indian Institute of Science Education and Research (IISER), Mohali}
\author{Arvind}
\email{arvind@iisermohali.ac.in}
\selectlanguage{british}%
\affiliation{Indian Institute of Science Education
and Research (IISER), Mohali}
\begin{abstract}
We propose a non-contextual hidden variable model,
consistent with all predictions of quantum
mechanics (QM).  A careful scrutiny of consistency
requirements of any hidden variable model with
quantum mechanics and the corresponding no-go
theorems, leads us to the conclusion that the
notion of contextuality is not a necessary feature
of QM.  The alternative view that emerges, hinges
on identifying a new classical property which we
call ``multiplicativity''.  It turns out that by
relaxing this condition of ``multiplicativity'' on
hidden variable models, they can be made
consistent with QM.  The connection of
non-multiplicativity to Bohm's model and to
apparent contextuality is also discussed.
\end{abstract}
\pacs{}
\maketitle
Quantum mechanics has been one of the most
successful theories in physics so far, however,
there has not yet been a final word on its
completeness and interpretation~\cite{BellSpkblUnspkbl}.
Einstein's~\cite{EinsteinEPR} work on the
incompleteness of QM and the subsequent seminal
work of Bell~\cite{BellSpkblUnspkbl}, assessing the
compatibility of a more complete model involving
hidden variables (HV) and locality with QM, has
provided deep insights into  how the quantum world
differs from its classical counterpart.  In recent
times, these insights have been of pragmatic
utility in the areas of quantum information
processing, where EPR pairs are fundamental motifs
of entanglement~\cite{Ekert,PironioRndmnssCrtfcn,NielsenChuang}.
The work of  Kochen Specker~\cite{KochenSpecker}
and Gleason et.
al.~\cite{Gleason,BellOnHiddenVariables,Peres,Mermin})
broadened the schism between  HV models and QM.
They showed that it was contextuality and not
non-locality which was at the heart of this schism
and the incompatibility between HV models and QM
can arise even for a single quantum system.
Contextuality has thus been identified as a
fundamental non-classical feature of the quantum
world and experiments have also been proposed and
conducted to this effect~\cite{SimonContExpProp,
HuangContExp}.
Contextuality, on the one hand has led to
investigations on the foundational aspects of
QM including attempts to prove completeness of QM~\cite{PawelCntxClsscl,CabelloMmryQM}, and on the other
hand has been harnessed for computation and
cryptography~\cite{HowardCntxCmptn,CabelloCntxScrt}.
However, not all HV
models are incompatible with QM, and a very
important case in point is  Bohm's model based on
precise trajectories for quantum
systems~\cite{Bohm1,Bohm2}.

A closer look at the aforesaid no-go theorems
reveals that contextuality is in fact not a
necessary feature of QM (see \figref{block}).
While this has been shown earlier by taking
disturbances due to measurements into
consideration~\cite{NoContextuality},
we take a different approach 
and construct a 
HV model whose 
predictions match with those of QM for systems
described by arbitrary finite-dimensional Hilbert
spaces.  \begin{figure}[h]
\includegraphics[width=0.9\columnwidth]{block1}
\caption{Non-contextuality is not inconsistent
with QM} \label{fig:block}\end{figure} The
algebraic constraints obeyed by quantum
observables may not necessarily be obeyed by the
predictions of an HV model at the level of
individual outcomes.  Those HV models that obey
such algebraic constraints possess a special
property which we call `multiplicativity'.  
Enlarging the scope of HV models to include
non-multiplicative models, permits us to construct an
explicit non-contextual model consistent with QM.
The re-examination of the very proof of
`contextuality'  is used to illustrate
non-multiplicativity, an intrinsic quantum feature, 
which we propose as a possible alternative to contextuality.

%{\noindent \bf Notation:} 
\begin{notn*}
(a) $\psi\in\mathcal{H}$ 
represents a pure quantum mechanical
state of the system in the Hilbert space
$\mathcal{H}$, (b) $\hat{\mathcal{H}}$ is defined
to mean the set of Hermitian 
observables for the system, (c)
$[\mathcal{H}]$ is defined to mean
$(\mathcal{H},\,\mathbb{R}^{\otimes})$, which
represents the state of the system including HVs,
(d) $[\psi]\in[\mathcal{H}]$ will represent the
state of the system, including HVs, (e) a
prediction map is
$M:\{ \hat{\mathcal{H}},[\mathcal{H}] \}\to\mathbb{R}$,
(f) a sequence map is
$S:\{ \hat{\mathcal{H}},[\mathcal{H}],\mathbb{R} \}\to[\mathcal{H}]$,
(g) $f$ is an arbitrary map from $\{
\hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}}
\} \to \hat{\mathcal{H}}$ constructed using
multiplication and addition of compatible observables,
and multiplication with complex
numbers
%\footnote{Strictly, the map $f$ would depend on the operators to ensure preservation of Hermiticity; thus a given $f$ will be defined only for a subset of $\{\hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}}\}$.}
, (h) $\tilde{f}$ is a map constructed by
replacing observables in $f$ with real numbers.
\end{notn*}

\begin{defn} A theory is non-contextual, if it
provides a map $M: \{
\hat{\mathcal{H}},[\mathcal{H}] \} \to\mathbb{R}$
to explain measurement outcomes. A theory which is
not non-contextual is contextual.\end{defn}

\begin{defnrem} The notion of non-contextuality may be
extended to maps. A map is non-contextual if it is
of the form $M: \{ \hat{\mathcal{H}},[\mathcal{H}]
\} \to\mathbb{R}$; prediction maps as defined above, for individual observables, are non-contextual by construction.  
\end{defnrem}
\begin{defnrem}
Broader definitions in the literature have been
suggested which declare a larger set of theories
as non-contextual. For our purposes, however, this
restricted definition will suffice. The term
contextual is used to suggest that the value an
operator takes might depend on which other
compatible observable it is being  measured with.
\end{defnrem}

Before we re-examine the result of Kochen and
Specker~\cite{KochenSpecker} that non-contextual
models cannot be compatible with quantum
mechanics and show that multiplicativity can provide an
alternative to contextuality, we define
multiplicativity more precisely.
\begin{defn}
A prediction map $M$ is \emph{multiplicative} iff
\begin{align*} &
M(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi])
= \\ & \tilde
f(M(\hat{B}_{1},[\psi]),M(\hat{B}_{2},[\psi]),\dots
M(\hat{B}_{N},[\psi])), \end{align*}
where $\hat{B}_{i}\in\hat{\mathcal{H}}$ are
arbitrary mutually commuting observables and
$[\psi]\in[\mathcal{H}]$. A
\emph{non-multiplicative} map is one that is not
multiplicative.  
\end{defn} 
Note that if $M$ is taken to represent the
measurement outcome (in QM), then for states of
the system which are simultaneous eigenkets of
$\hat{B}_{i}$s, $M$ must clearly be
multiplicative. It is, however, not obvious that
this property must always hold. For example,
consider two spin-half particles in the state 
$\left|1\right\rangle \otimes\left|1\right\rangle
$ written in the computational basis and take
operators 
$\hat{B}_{1}=\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}$,
$\hat{B}_{2}=\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}$
and
$\hat{C}=\hat{B}_{1}\hat{B}_{2}=-\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}$
written in terms of Pauli operators.
We must have $M(\hat{C})=-1$ while
$M(\hat{B}_{1})=\pm1$ and $M(\hat{B}_{2})=\pm1$
independently, according to QM, with probability
half. Here multiplicativity does not seem to hold.
Antithetically, it is clear that if one first
measures $\hat{B}_{1}$ and subsequently measures
$\hat{B}_{2}$, then the product of the results
must be $-1$. This is consistent with measuring
$\hat{C}$. To capture this property we define
\emph{sequential multiplicativity} as follows.
\begin{defn} 
A prediction map $m$ is
\emph{sequentially multiplicative} for a given
sequence map $s$, iff \begin{align*}
&M(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi_{1}])=\\
&\tilde
f(M(\hat{B}_{1},[\psi_{k_{1}}]),M(\hat{B}_{2},[\psi_{k_{2}}]),\dots,M(\hat{B}_{N},[\psi_{k_{N}}])),
\end{align*} 
where
$\mathbf{k}=(k_{1},k_{2},\dots,k_{N})\in\{
N! {\rm ~permutations~of~}  k'{\rm s}
\}$, $\hat{B}_{i}\in\hat{\mathcal{H}}$ are
arbitrary mutually commuting observables,
$[\psi_{i}]\in[\mathcal{H}]$ and
$[\psi_{k+1}] :=
S(\hat{B}_{k},[\psi_{k}],M(\hat{B}_{k},[\psi_{k}]))$,
$\forall\,[\psi_{i}]$.  
\label{defn:seqnMltpl}\end{defn} 
With these definitions we are now ready to discuss
the `proof of contextuality'. We first state the
contextuality theorem in our notation:
\begin{thm*} Let a map
$M:\hat{\mathcal{H}}\to\mathbb{R}$, be s.t. (a)
$M(\hat{\mathbb{I}})=1$, (b)
$M(f(\hat{B}_{1},\hat{B}_{2},\dots))=\tilde
f(M(\hat{B}_{1}),M(\hat{B}_{2}),\dots)$, for any
arbitrary function $f$, where $\hat{B}_{i}$ are
mutually commuting Hermitian operators. If $m$ is
assumed to describe the outcomes of measurements,
then no $M$ exists which is consistent with all
predictions of QM. 
\label{thm:KS}
\end{thm*}

\begin{proof} Peres Mermin [PM]
($\left|\mathcal{H}\right|\ge4$)
\cite{Peres,Mermin}:
For a system composed of two spin-half particles 
consider the following set of operators \[
\hat{A}_{ij}\doteq\left[\begin{array}{ccc}
\hat{\mathbb{I}}\otimes\hat{\sigma}_{x} &
\hat{\sigma}_{x}\otimes\hat{\mathbb{I}} &
\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}\\
\hat{\sigma}_{y}\otimes\hat{\mathbb{I}} &
\hat{\mathbb{I}}\otimes\hat{\sigma}_{y} &
\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}\\
\hat{\sigma}_{y}\otimes\hat{\sigma}_{x} &
\hat{\sigma}_{x}\otimes\hat{\sigma}_{y} &
\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}
\end{array}\right] \] which have the property that
all operators along a row (or column) commute. Further,
the product of rows (or columns) yields
$\hat{R}_{i}=\hat{\mathbb{I}}$ and
$\hat{C}_{j}=\hat{\mathbb{I}}\,(j\neq3)$,
$\hat{C}_{3}=-\hat{\mathbb{I}}$, ($\forall\,i,j$) where
$\hat{R}_{i}:=\prod_{j}\hat{A}_{ij}$,
$\hat{C}_{j}:=\prod_{i}\hat{A}_{ij}$. Let us
assume that $M$ exists. From property (b) of the
map, to get $M(\hat{C}_{3})=-1$ (as required by
property (a)), we must have an odd number of $-1$
assignments in the third column. In the remaining
columns, the number of $-1$ assignments must be
even (for each column). Thus, in the entire
square, the number of $-1$ assignments must be
odd. Let us use the same reasoning, but along the
rows. Since each $M(\hat{R}_{i})=1$, we must have
even number of $-1$ assignments along each row.
Thus, in the entire square, the number of $-1$
assignments must be even. We have arrived at a
contradiction and therefore we conclude that our
assumption that a consistent $M$ exists, must be
wrong.  \end{proof}

\begin{rem} One could in principle assume $M$, to
be s.t. (a) $M(\hat{\mathbb{I}})=1$, (b)
$M(\alpha\hat{B}_{i})=\alpha M(\hat{B}_{i})$, for
$\alpha\in\mathbb{R}$, (c)
$M(\hat{B}_{i}^{2})=M(\hat{B}_{i})^{2}$, (d)
$M(\hat{B}_{i}+\hat{B}_{j})=M(\hat{B}_{i})+M(\hat{B}_{j})$,
to deduce (d')
$M(\hat{B}_{i}\hat{B}_{j})=M(\hat{B}_{i})M(\hat{B}_{j})$
and that $M(\hat{B}_{i})\in$ spectrum of
$\hat{B}_{i}$. Effectively then, condition (b)
listed in the theorem is satisfied as a
consequence.  Therefore, assuming (a)-(d) as
listed here, rules out a larger class of $M$~\cite{KochenSpecker}.
\end{rem} 
Here $M$ maybe viewed as a specific
class of prediction maps, that implicitly depends
on the state $[\psi]$. It is clear that according to
the theorem, non-contextual maps which are
\emph{multiplicative} must be incompatible with
QM. However, non-contextual maps which are
non-\emph{multiplicative} might be consistent with
QM. Indeed it is and we will show this by constructing an
explicit model.

>>>>

\documentclass[british,aps,prl,superscriptaddress,nofootinbib,times,reprint]{revtex4-1}
%\documentclass[british,aps,superscriptaddress,nofootinbib,twocolumn,prl]{revtex4-1}
%\usepackage[T1]{fontenc}
%\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm,headheight=2cm,headsep=2cm,footskip=2cm}
\setcounter{secnumdepth}{3}
\usepackage{verbatim}
\usepackage{refstyle}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
%\usepackage{esint}
\makeatletter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
\AtBeginDocument{\providecommand\thmref[1]{\ref{thm:#1}}}
\AtBeginDocument{\providecommand\lemref[1]{\ref{lem:#1}}}
\AtBeginDocument{\providecommand\propref[1]{\ref{prop:#1}}}
\AtBeginDocument{\providecommand\eqref[1]{\ref{eq:#1}}}
\AtBeginDocument{\providecommand\figref[1]{\ref{fig:#1}}}
\AtBeginDocument{\providecommand\secref[1]{\ref{sec:#1}}}
\AtBeginDocument{\providecommand\ssecref[1]{\ref{ssec:#1}}}
\AtBeginDocument{\providecommand\defnref[1]{\ref{defn:#1}}}
\AtBeginDocument{\providecommand\tblref[1]{\ref{tbl:#1}}}

@@

\begin{document}
\title{A non-contextual hidden variable model for quantum mechanics}
\author{Atul Singh Arora}
\email{ms11003@iisermohali.ac.in}
%\selectlanguage{british}%
\affiliation{Department of Physical Sciences, 
Indian Institute of Science Education and Research
Mohali, Sector 81 SAS Nagar 140306 Punjab India.}
\author{Arvind}
\email{arvind@iisermohali.ac.in}
\affiliation{Department of Physical Sciences, 
Indian Institute of Science Education and Research
Mohali, Sector 81 SAS Nagar 140306 Punjab India.}
%\selectlanguage{british}%
\begin{abstract}
We propose a non-contextual hidden variable model,
consistent with all {\color{red} kinematic
predictions of} quantum
mechanics (QM).  A careful scrutiny of consistency
requirements of any hidden variable model with
quantum mechanics and the corresponding no-go
theorems leads us to the conclusion that,
contrary to common belief,  the notion of
contextuality is not a necessary feature of QM.
The alternative view that emerges, hinges on
identifying a new classical property which we call
``multiplicativity''.  It turns out that by
relaxing this condition of ``multiplicativity'' on
hidden variable models, they can be made
consistent with QM  while remaining
non-contextual.  The connection of
non-``multiplicativity'' to Bohm's model and to
apparent contextuality is also discussed.
\end{abstract}
\pacs{}
\maketitle
Quantum mechanics has been one of the most
successful theories in physics so far, however,
there has not yet been a final word on its
completeness and
interpretation~\cite{BellSpkblUnspkbl}.
Einstein's~\cite{EinsteinEPR} work on the
incompleteness of QM and the subsequent seminal
work of Bell~\cite{BellSpkblUnspkbl}, assessing
the compatibility of a more complete model
involving hidden variables (HV) and locality with
QM, has provided deep insights into  how the
quantum world differs from its classical
counterpart.  In recent times, these insights have
been of pragmatic utility in the area of quantum
information processing (QIP), where EPR pairs are
fundamental motifs of
entanglement~\cite{Ekert,PironioRndmnssCrtfcn,NielsenChuang}.
The work of  Kochen Specker~\cite{KochenSpecker}
and Gleason {\it
et.\,al.}~\cite{Gleason,Peres,Mermin} broadened
the schism between  HV models and QM.  They showed
that it was contextuality and not non-locality
which was at the heart of this schism and the
incompatibility between HV models and QM can arise
even for a single quantum system.  Contextuality
has thus been identified as a fundamental
non-classical feature of the quantum world and
experiments have also been proposed and conducted
to this effect~\cite{SimonContExpProp,
HuangContExp}.
Contextuality, on the one hand has
led to investigations on the foundational aspects
of QM including attempts to prove completeness of
QM~\cite{PawelCntxClsscl,CabelloMmryQM}, and on
the other hand has been harnessed for computation
and
cryptography~\cite{HowardCntxCmptn,CabelloCntxScrt}.
However, not all HV models (e.g.
Bohm's model based on
trajectories~\cite{Bohm1,Bohm2}) are incompatible
with QM~\cite{BellOnHiddenVariables}, and we in
this letter present a new non-contextual HV model
consistent with QM.
\begin{figure}[h]
\includegraphics[width=0.9\columnwidth]{block1}
\caption{Non-contextuality is not inconsistent
with QM} \label{fig:block}\end{figure}
A closer scrutiny of the arguments
leading to the dichotomy between QM and HV models
reveals that the algebraic constraints obeyed by
quantum observables may not necessarily be obeyed
by the predictions of an HV model at the level of
individual outcomes and  that contextuality is 
not a necessary feature of QM.  Those HV
models that obey such algebraic constraints
possess a special property which we call
``multiplicativity''.  Enlarging the scope of HV
models to include non-multiplicative models,
permits us to construct an explicit non-contextual
HV model consistent with 
{\color{red} kinematics predictions of}
QM (see \figref{block}).
The re-examination of the very proof of
`contextuality'  is used to illustrate
non-multiplicativity, an intrinsic quantum
feature, which we propose as a possible
alternative to contextuality. 
Non-mulitiplicativity thus might turn out to be
the key quantum feature distinguishing the quantum
world from the classical one and may have a role
to play in QIP protocols.  This provides a new
way to look at the classical-quantum divide and at
the foundations of quantum mechanics.

\begin{notn*}
(a) $\psi\in\mathcal{H}$ represents a pure quantum
mechanical state of the system in the Hilbert
space $\mathcal{H}$, (b) $\hat{\mathcal{H}}$ is
defined to mean the set of Hermitian observables
for the system, (c) $[\mathcal{H}]$ is defined to
mean $\{\mathcal{H},\,\mathbb{R}^{\otimes}\}$, which
represents the state  space of the
system including HVs, (d) $[\psi]\in[\mathcal{H}]$
will represent the state of the system including
HVs, (e) a prediction map is $M:\{
\hat{\mathcal{H}},[\mathcal{H}] \}\to\mathbb{R}$,
(f) a sequence map is $S:\{
\hat{\mathcal{H}},[\mathcal{H}],\mathbb{R}
\}\to[\mathcal{H}]$, (g) $f$ is an arbitrary map
from $\{
\hat{\mathcal{H}},\hat{\mathcal{H}},\dots\hat{\mathcal{H}}
\} \to \hat{\mathcal{H}}$ constructed using
multiplication and addition of compatible
observables, and multiplication with complex
numbers, (h) $\tilde{f}$ is a map constructed by
replacing observables in $f$ with real numbers.
\end{notn*}
\begin{defn} A theory is non-contextual, if it
provides a map $M: \{
\hat{\mathcal{H}},[\mathcal{H}] \} \to\mathbb{R}$
to explain measurement outcomes. A theory which is
not non-contextual is contextual~\cite{peresBook}.
\end{defn}
{\color{red}
\begin{defnrem}
A prediction map of the form $M: \{
\hat{\mathcal{H}},[\mathcal{H}] \} \to\mathbb{R}$.
itself can be called  non-contextual.
\end{defnrem}
}
\begin{defnrem}
Broader definitions in the literature have been
suggested which declare a larger set of theories
as non-contextual~\cite{spekkens_pra}. For our
purposes, however, this restricted definition will
suffice  and our results will in any case be valid
for the broader definitions. The term contextual
is used to suggest that the value an observable
takes might depend on which other compatible
observable it is being  measured with.
\end{defnrem}
Before we re-examine the result of Kochen and
Specker~\cite{KochenSpecker} that non-contextual
models cannot be compatible with quantum
mechanics and show that multiplicativity can provide an
alternative to contextuality, we define
multiplicativity explicitly.
\begin{defn}
A prediction map $M$ is \emph{multiplicative} iff
\begin{align*} &
M(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi])
= \\ & \tilde
f(M(\hat{B}_{1},[\psi]),M(\hat{B}_{2},[\psi]),\dots
M(\hat{B}_{N},[\psi])), \end{align*}
where $\hat{B}_{i}\in\hat{\mathcal{H}}$ are
arbitrary mutually commuting observables and
$[\psi]\in[\mathcal{H}]$. A
\emph{non-multiplicative} map is one that is not
multiplicative.  
\end{defn} 
Note that if $M$ is taken to represent the
measurement outcome (in QM), then for states of
the system which are simultaneous eigenkets of
$\hat{B}_{i}$s, $M$ must clearly be
multiplicative. It is, however, not obvious that
this property must always hold. For example,
consider two spin-half particles in the state 
$\left|1\right\rangle \otimes\left|1\right\rangle
$ written in the computational basis and 
operators 
$\hat{B}_{1}=\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}$,
$\hat{B}_{2}=\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}$
and
$\hat{C}=\hat{B}_{1}\hat{B}_{2}=-\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}$
written in terms of Pauli operators.
We must have $M(\hat{C})=-1$ while
$M(\hat{B}_{1})=\pm1$ and $M(\hat{B}_{2})=\pm1$
independently, according to QM, with probability
half. Here multiplicativity clearly is not
required to hold.
Antithetically, it is clear that if one first
measures $\hat{B}_{1}$ and subsequently measures
$\hat{B}_{2}$, then the product of the results
must be $-1$. This is consistent with measuring
$\hat{C}$. To capture this property we define
\emph{sequential multiplicativity} as follows.
\begin{defn} 
A prediction map $m$ is
\emph{sequentially multiplicative} for a given
sequence map $s$, iff \begin{align*}
&M(f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{N}),[\psi_{1}])=\\
&\tilde
f(M(\hat{B}_{1},[\psi_{k_{1}}]),M(\hat{B}_{2},[\psi_{k_{2}}]),\dots,M(\hat{B}_{N},[\psi_{k_{N}}])),
\end{align*} 
where
$\{k_{1},k_{2},\dots,k_{N}\} \in\{
N! {\rm ~permutations~of~}  k'{\rm s} \}$, 
$\hat{B}_{i}\in\hat{\mathcal{H}}$ are
arbitrary mutually commuting observables,
$[\psi_{i}]\in[\mathcal{H}]$ and
$[\psi_{k+1}] :=
S(\hat{B}_{k},[\psi_{k}],M(\hat{B}_{k},[\psi_{k}]))$,
$\forall\,[\psi_{i}]$.  
\label{defn:seqnMltpl}
\end{defn} 
With these definitions we are now ready to discuss
the `proof of contextuality'. We first state the
contextuality theorem in our notation:
\begin{thm*} Let a map
$M:\hat{\mathcal{H}}\to\mathbb{R}$, be s.t. (a)
$M(\hat{\mathbb{I}})=1$, (b)
$M(f(\hat{B}_{1},\hat{B}_{2},\dots))=\tilde
f(M(\hat{B}_{1}),M(\hat{B}_{2}),\dots)$, for any
arbitrary function $f$, where $\hat{B}_{i}$ are
mutually commuting Hermitian operators. If $m$ is
assumed to describe the outcomes of measurements,
then no $M$ exists which is consistent with all
predictions of QM. 
\label{thm:KS}
\end{thm*}

\begin{proof} Peres Mermin [PM]
($\left|\mathcal{H}\right|\ge4$)
\cite{Peres,Mermin}:
For a system composed of two spin-half particles 
consider the following set of operators \[
\hat{A}_{ij}\doteq\left[\begin{array}{ccc}
\hat{\mathbb{I}}\otimes\hat{\sigma}_{x} &
\hat{\sigma}_{x}\otimes\hat{\mathbb{I}} &
\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}\\
\hat{\sigma}_{y}\otimes\hat{\mathbb{I}} &
\hat{\mathbb{I}}\otimes\hat{\sigma}_{y} &
\hat{\sigma}_{y}\otimes\hat{\sigma}_{y}\\
\hat{\sigma}_{y}\otimes\hat{\sigma}_{x} &
\hat{\sigma}_{x}\otimes\hat{\sigma}_{y} &
\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}
\end{array}\right] \] which have the property that
all operators along a row (or column) commute. Further,
the product of rows (or columns) yields
$\hat{R}_{i}=\hat{\mathbb{I}}$ and
$\hat{C}_{j}=\hat{\mathbb{I}}\,(j\neq3)$,
$\hat{C}_{3}=-\hat{\mathbb{I}}$, ($\forall\,i,j$) where
$\hat{R}_{i}:=\prod_{j}\hat{A}_{ij}$,
$\hat{C}_{j}:=\prod_{i}\hat{A}_{ij}$. Let us
assume that $M$ exists. From property (b) of the
map, to get $M(\hat{C}_{3})=-1$ (as required by
property (a)), we must have an odd number of $-1$
assignments in the third column. In the remaining
columns, the number of $-1$ assignments must be
even (for each column). Thus, in the entire
square, the number of $-1$ assignments must be
odd. Let us use the same reasoning, but along the
rows. Since each $M(\hat{R}_{i})=1$, we must have
even number of $-1$ assignments along each row.
Thus, in the entire square, the number of $-1$
assignments must be even. We have arrived at a
contradiction and therefore we conclude 
that $M$ does not exist
\end{proof}

\begin{rem} One could in principle assume $M$, to
be s.t. (a) $M(\hat{\mathbb{I}})=1$, (b)
$M(\alpha\hat{B}_{i})=\alpha M(\hat{B}_{i})$, for
$\alpha\in\mathbb{R}$, (c)
$M(\hat{B}_{i}^{2})=M(\hat{B}_{i})^{2}$, (d)
$M(\hat{B}_{i}+\hat{B}_{j})=M(\hat{B}_{i})+M(\hat{B}_{j})$,
to deduce (d)
$M(\hat{B}_{i}\hat{B}_{j})=M(\hat{B}_{i})M(\hat{B}_{j})$
and that $M(\hat{B}_{i})\in$ spectrum of
$\hat{B}_{i}$. Effectively then, condition (b)
listed in the theorem is satisfied as a
consequence.  Therefore, assuming (a)-(d) as
listed above, rules out a larger class of
$M$~\cite{KochenSpecker}.
\end{rem} 
Here $M$ maybe viewed as a specific class of
prediction maps, that implicitly depends on the
state $[\psi]$. It is clear that according to the
theorem, non-contextual maps which are
\emph{multiplicative} must be incompatible with
QM. 
This leaves open an interesting possibility
that non-contextual maps which are
non-\emph{multiplicative} could be consistent with
QM. We explicitly construct such an HV model and use
\emph{sequential multiplicativity} as a consistency
check imposed by QM.
\begin{prop*} 
Let a quantum mechanical 
system be in a state, s.t. measurement of
$\hat{C}$ yields repeatable results (same result
each time). Then according to QM, \emph{sequential
multiplicativity} holds, where $\hat{C}:=
f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n})$, and
$\hat{B}_{i}$ are as defined (in \defnref{seqnMltpl})
\end{prop*}
\begin{proof} 
Without loss of generality we can take
$\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n}$ to be
mutually compatible and a complete set of
observables (operators can be added to make the set
complete if needed).
It follows
that $\exists$
$\left|\mathbf{b}=\left(b_{1},b_{2},\dots
b_{n}\right)\right\rangle $ s.t.
$\hat{B}_{i}\left|\mathbf{b}\right\rangle
=b_{i}\left|\mathbf{b}\right\rangle $,
and that
$\sum_{\mathbf{b}}\left|\mathbf{b}\right\rangle
\left\langle \mathbf{b}\right|=\hat{\mathbb{I}}$.
Let the state of the system
$\left|\psi\right\rangle $ be s.t.
$\hat{C}\left|\psi\right\rangle
=c\left|\psi\right\rangle $. For
the statement to follow, one need only show that
$\left|\psi\right\rangle $ must be made of only
those $\left|\mathbf{b}\right\rangle $s, which
satisfy $c=\tilde
f(b_{1},b_{2},\dots
b_{n})$.  This is the crucial step and
proving this is straightforward. We start with
$\hat{C}\left|\psi\right\rangle
=c\left|\psi\right\rangle $ and take its inner
product with $\left\langle \mathbf{b}\right|$ to
get 
\begin{eqnarray*} 
c \langle \mathbf{b}|\psi\rangle 
& = &
\langle
\mathbf{b} |\hat{C} |\psi
\rangle  
\\
&=&
\langle
\mathbf{b}| f(\hat{B}_{1},\hat{B}_{2},\dots\hat{B}_{n}) |
\psi \rangle
\\ 
&=&
\tilde f (b_{1},b_{2},\dots b_{n})
\langle
\mathbf{b}|\psi\rangle  
\end{eqnarray*}
Also, we have $\left|\psi\right\rangle
=\sum_{\mathbf{b}}\left\langle
\mathbf{b}|\psi\right\rangle
\left|\mathbf{b}\right\rangle $, from
completeness. If we consider
$\left|\mathbf{b}\right\rangle $s for which
$\left\langle \mathbf{b}|\psi\right\rangle \neq0$,
then we find that indeed $c=\tilde
f(b_{1},b_{2},\dots
b_{n})$. The case when $\left\langle
\mathbf{b}|\psi\right\rangle =0$ is anyway
irrelevant as the corresponding $\vert b\rangle$
does not contribute to $\vert \psi\rangle$.
We can thus conclude that
$\left|\psi\right\rangle $ is made only of those
$\left|\mathbf{b}\right\rangle $s that satisfy the
required relation.
\end{proof} 
It is worth noting that in the Peres Mermin
case, where $\hat{R}_{i}$ and $\hat{C}_{j}$ are
$\pm\hat{\mathbb{I}}$, it follows that all
states are their eigenstates. Consequently, for
these operators \emph{sequential multiplicativity}
must always hold.
%\subsection{C-ingle model | An Explicit
%Construction \label{ssec:cingle}}

We are now ready to describe our 
model explicitly.
Let the state of a finite dimensional quantum
system be $\left|\chi\right\rangle $. We wish to
assign a value to an arbitrary observable
$\hat{A}=\sum_{a}a\left|a\right\rangle
\left\langle a\right|$, which has eigenvectors $\{
\vert a_j\rangle \}$. The corresponding ordered
eigenvalues are $\{a_j\}$ such that $a_{\rm
min}=a_1$ and $a_{\rm max}=a_n$.  Our HV model for
QM assigns values in the following three steps:
\setdefaultleftmargin{0pt}{}{}{}{}{}
\begin{enumerate}
\item
{\bf  Initial HV:} Pick a number
$c\in[0,1]$, from a uniform random distribution.\
\item
{\bf
Assignment or Prediction:}
 The value assigned to
$\hat{A}$ is given by finding the smallest $a$
s.t.  $c\le\sum_{a'=a_{\text{min}}}^{a}\left|\left\langle
a'|\chi\right\rangle \right|^{2},$ viz. we have
specified a prediction map, $M(\hat{A},[\chi ])=a$.
\item
{\bf  Update:} After measuring an operator, the state is
updated (collapsed) in accordance with the rules
of QM. This completely specifies the sequence map
$S$.
\end{enumerate}
The above HV model works for all quantum systems,
however, to illustrate its working,
consider the example of a spin-half particle in the state
$\left|\chi\right\rangle
=\cos\theta\left|0\right\rangle
+\sin\theta\left|1\right\rangle $ and the observable
$\hat{A}=\hat{\sigma}_{z}=\left|0\right\rangle
\left\langle 0\right|-\left|1\right\rangle
\left\langle 1\right|$. Now, according to the
postulates of this theory, $M(\hat{A},[\chi])=+1$
if
the randomly generated value for 
$c\le\cos^{2}\theta$ else $\hat{A}$ is assigned
$-1$; it then follows,  from $c$ being uniformly
random in $[0,1]$ that the statistics agree with
the predictions of QM {\it i.e.}  the
Born rule. 
The assignment described by the
prediction map $M$ is non-contextual since, given
an operator and a state (alongwith the HV $c$), the value is
uniquely assigned. The map $M$ is, however,
non-multiplicative.
